ML Estimation of Matrix--$t$ Distributions with Expectation Maximization

### 2017-01-22 paper.tex
Arranged the sections.

This note is organized as follows. In section \ref{sec:derivation}, we present an EM formulae for MLE of matrix--$t$ distribution together with the relevant details on the derivation. The notation and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section. Practical implementation of the algorithm is discussed in section \ref{sec:implement} and a brief conclusion is given in section \ref{sec:conclusion}.

\section{EM for matrix--$t$ distribution}\label{sec:derivation}

All matrices referred in this note are real. For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_+^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$. The multivariate gamma
function, $\Gamma_{p}(\cdot)$, given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
\begin{equation}
\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
We also define the multivariate digamma function
%
\begin{equation}
\psi_{p}(x):=\frac{\partial}{\partial x}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
\end{equation}

\subsection{Matrix--$t$ distribution}

@@

\subsection{EM formulae for matrix--$t$ distribution}

@@

@@

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
\end{split}
\label{eq:fsimp}
\end{equation}
where the matrices $W_j^{(k)}$ are defined as
\begin{equation}
W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).
\end{equation}
For the rest of the parameter(s), we consider two possible situations.
\begin{itemize}
\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
\begin{equation}
\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:sigma}
\end{equation}
$\Sigma_{k+1}$ has linear dependence on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
is a concave function of $\nu$ for all $\nu>0$. The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive. For this case we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$, which has the form
\begin{equation}
h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0
\end{equation}
with $h(x):=p\log(x)-\psi_p(x/2)$.
\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$. The gradient equation of $\nu$ is given by
\begin{equation}
\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.
\end{equation}
\end{itemize}

\section{Implementation}\label{sec:implement}
In the actual implementation of

>>>>

This note is organized as follows. In section \ref{sec:matrixt} we give a brief review on matrix--$t$ distributions. The notations and the preliminary results on matrix calculus needed to present the derivation are presented sequentially in the section. EM formulae together with the relevant derivation and implementation details are provided in section \ref{sec:EM} and a brief conclusion is given in section \ref{sec:conclusion}.

\section{Matrix--$t$ distributions}\label{sec:matrixt}

All matrices referred in this note are real. For notation convenience, the matrix determinant is denoted as $|\cdot|$ and $\mathbf{S}_+^p$ is used to indicate the set of symmetric positive definite (SPD) matrices of dimension $p$. The multivariate gamma function $\Gamma_{p}(\cdot)$ given in terms of the standard gamma function $\Gamma(\cdot)$ has the form
\begin{equation}
\Gamma_{p}(x)=\pi^{p(p-1)/4}\prod_{j=1}^{p}\Gamma\left(x+\frac{1-j}{2}\right).\label{eq:multgammadef}
\end{equation}
We also define the multivariate digamma function
%
\begin{equation}
\psi_{p}(x):=\frac{d}{dx}\log\Gamma_{p}(x)=\sum_{j=1}^{p}\psi\left(x+\frac{1-j}{2}\right).
\end{equation}

@@

\section{EM formulae for matrix--$t$ distribution}\label{sec:EM}

@@

\subsection{Expectation step}

@@

Utilizing the equalities \eqref{eq:intderivative1} and \eqref{eq:intderivative2} with adequate rearrangements we obtain the closed form expression of $Q_j(\theta,\theta_k)$ while ignoring the irrelevant constant terms
\begin{equation}
\begin{split}Q_j(\theta,\theta_k) \sim \nu\left\{\log|W_j^{(k)}|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)\right\}-2\log\Gamma_{p}\left(\frac{\nu+p-1}{2}\right)-p\log|\Omega|\\
-(m+\nu_k+p-1)\textrm{tr}\left[W_j^{(k)}\left\{\Sigma+(T_j-M)\Omega^{-1}(T_j-M)^{\textrm{t}}\right\}\right]+(\nu+p-1)\log |\Sigma|,
\end{split}
\label{eq:fsimp}
\end{equation}
where the matrices $\left\{W_j^{(k)}\right\}_{j\in J}$ are defined as
\begin{equation}
W_j^{(k)}:=\left\{\Sigma_k+(T_j-M_k)\Omega_k^{-1}(T_j-M_k)^{\textrm{t}}\right\}^{-1}.
\end{equation}

\subsection{Maximization step}

To maximize $Q(\theta,\theta_k)$ with respect to $\theta$ we compute the gradient of each parameter and equate them to $0$ to obtain $\theta_{k+1}$. For parameter $M$ we have
\begin{equation}
\sum_{j\in J}W_j^{(k)}\left(T_j-M\right)\Omega^{-1}=0.
\end{equation}
As $\Omega$ is nonsingular we obtain $M_{k+1}$ as an average of the input data weighted by $W_j^{(k)}$
\begin{equation}
M_{k+1}=\left(\sum_{j\in J}W_j^{(k)}\right)^{-1}\sum_{j\in J}W_j^{(k)}T_j.\label{eq:M}
\end{equation}
Maximizing $Q(\theta,\theta_k)$ with respect to $\Omega$ we obtain $\Omega_{k+1}$
\begin{equation}
\Omega_{k+1}=\frac{m+\nu_k+p-1}{p}\left(\frac{1}{|J|}\sum_{j\in J}(T_j-M_{k+1})^{\textrm{t}}W_j^{(k)}(T_j-M_{k+1})\right).\label{eq:Omega}
\end{equation}
For the rest of the parameter(s), we consider two possible situations.
\begin{itemize}
\item $\Sigma$ and $\nu$ are unknown: In this situation, maximizing $Q(\theta,\theta_k)$ with respect to $\Sigma$ we obtain $\Sigma_{k+1}$ in terms of the inverse of the mean of $W_j^{{(k)}}$
\begin{equation}
\Sigma_{k+1}(\nu)=\frac{\nu+p-1}{m+\nu_k+p-1}\left(\bar{W}^{(k)}\right)^{-1},\quad \bar{W}^{(k)}:=\frac{1}{|J|}\sum_{j\in J}W_j^{(k)}\label{eq:Sigma}
\end{equation}
$\Sigma_{k+1}$ has linear dependence on $\nu$. Due to this, we substitute $\Sigma_{k+1}$ back into $Q(\theta,\theta_k)$ and maximize it with repect to $\nu$. We remark that
\[\tilde{Q}(\nu):=Q\left(\left\{M_{k+1},\Sigma_{k+1}(\nu),\Omega_{k+1},\nu\right\},\theta_k\right)\]
is a concave function of $\nu$ for all $\nu>0$. The maximum only attains at positive $\nu$ if $\tilde{Q}'(0)$ is positive. For this case we obtain $\nu_{k+1}$ by solving the equation $\tilde{Q}'(\nu)=0$, which has the form
\begin{equation}
h(\nu+p-1)-h(m+\nu_k+p-1)+\frac{1}{|J|}\sum_{j\in J}\log|W_j^{(k)}|-\log\left|\bar{W}^{(k)}\right|=0\label{eq:nu1}
\end{equation}
with $h(x):=p\log(x)-\psi_p(x/2)$.
\item $\Sigma$ is known and $\nu$ is unknown: In this situation, $Q\left(\left\{M_{k+1},\Sigma,\Omega_{k+1},\nu\right\},\theta_k\right)$ is also a concave function of $\nu$. The gradient equation of $\nu$ is given by
\begin{equation}
\log|\Sigma|-\log\left|\bar{W}^{(k)}\right|+\psi_{p}\left(\frac{m+\nu_k+p-1}{2}\right)-\psi_{p}\left(\frac{\nu+p-1}{2}\right)=0.\label{eq:nu2}
\end{equation}
\end{itemize}
